{\Huge Supersymmetry} \\ \vspace{5mm} Lecture notes for FYS5190/FYS9190 Supersymmetry

### 2025-11-17 mssm.tex
Small changes to MSSM.tex

With all possible apologies, we now change notation for the component fields to what is closer to conventions in phenomenology, as opposed to pure theory. The left-handed Weyl spinors in these superfields will now be named for example $l_i$ for the spinor in the superfield $l_i$, $\nu_i$ for the one in the $\nu_i$ superfield, and $\bar e_{i}$ for the one in $\bar E_i$,  $u_{i}$ for the one in the $u_i$ superfield, $d_i$ for the one in the $d_i$ superfield, $\bar u_i$ for the one in the $\bar U_i$ superfield,  and finally $\bar d_i$ for the one in the $\bar D_i$ superfield. This means that we leave our former notation where the bar signifies right-handed Weyl spinor, but now instead signifies the left-handed Weyl spinor for the anti-particle, or, in other words, a $SU(2)_L$ singlet. If we now want to indicate a right-handed Weyl spinor we use the hermitian conjugation in (\ref{eq:Weyl_hc}) as an alternative notation, so that for example the $l_i^\dagger$ superfield contains $\bar l_i^\dagger$ and $\bar E_i^\dagger$ contains $\bar e_i^\dagger$, and so on. The reason for this change is in part that we need a simple way to write down component field Lagrangians with fermions without greek letters and very many indices indicating particle type, as well as running out of letters since we need two superfields for one fermion.
% An alternative notation would be to write $e_{Li}$ for the left-handed Weyl-spinor in $l_i$ and $e_{Ri}$ for the left-handed Weyl-spinor in $\bar E_i$ etc.
% TODO: Work out the Weyl -> Dirac transition and re-consider the notation

@@

Looking at the construction $V\equiv qT_aV^a$ for the vector superfields in the supersymmetric Lagrangian we see that, as expected, we need one superfield $V^a$ per generator $T_a$ of the algebra, with the normal $SU(3)_c$, $SU(2)_L$ and $U(1)_Y$ vector bosons as vector component fields. We call these superfields $C^a$, $W^i$ and $B^0$, where $a=1,\ldots,8$ and $i=1,2,3$.\footnote{And there we have another W.} 
In order to be really confusing, we use the following symbols for the fermion Weyl-spinors: $\tilde g^a$, $\tilde W^+$, $\tilde W^0$, $\tilde W^-$ and $\tilde B^0$. The tilde here is supposed to tells us -- just as for the scalar component fields of the scalar superfields -- that they are supersymmetric partners of the known Standard Model particles. These particles are collectively known as the {\bf gauginos}.

@@

%%%%%%%%%%%%%
\section{The kinetic terms}
\label{sec:mssm_kinetic_terms}
%%%%%%%%%%%%%
It is now straight forward to write down the kinetic terms of the MSSM Lagrangian giving the matter-gauge interaction terms
\begin{eqnarray}\label{eq:kinlag}
\mathcal{L}_\text{kin} &=& L_i^\dagger e^{\frac{1}{2}g \sigma W - \frac{1}{2} g' B}L_i + Q_i^\dagger e^{\frac{1}{2}g_s\lambda C + \frac{1}{2}g\sigma W + \frac{1}{3}\cdot \frac{1}{2} g' B}Q_i \nonumber\\
&&+\bar{U}_{i}^\dagger e^{\frac{1}{2}g_s \lambda C - \frac{4}{3}\cdot \frac{1}{2} g' B}\bar{U}_i + \bar{D}_i^\dagger e^{\frac{1}{2} g_s\lambda C + \frac{2}{3}\cdot \frac{1}{2} g' B}\bar{D}_i \nonumber\\
&& + \bar{E}_i^\dagger e^{2\frac{1}{2} g' B}\bar{E}_i + H_u^\dagger e^{\frac{1}{2} g \sigma W + \frac{1}{2} g'B}H_u + H_d^\dagger e^{\frac{1}{2}g \sigma W - \frac{1}{2} g' B}H_d,
\end{eqnarray}
where $g'$, $g$ and $g_s$ are the couplings constants (strengths) of $U(1)_Y$, $SU(2)_L$, and $SU(3)_c$, respectively, and $\half\sigma_i$ and $\half\lambda_a$ are the generators of  $SU(2)_L$ and $SU(3)_c$. As a convention we assign the charges under $U(1)$, hypercharge, in units of $\frac{1}{2}g'$. For a list of the hypercharge assignments see Table~\ref{tab:hyper}.  All non-singlets of $SU(2)_L$ and $SU(3)_C$ have the same charge, the factor $\frac{1}{2}$ here is used to get by without accumulation of numerical factors since the algebras for the Pauli ($\sigma_i$) and Gell-Mann matrices ($\lambda_a$) are:
\[ \left[\frac{1}{2}\sigma_i, \frac{1}{2} \sigma_j\right] = i \epsilon_{ijk}\frac{1}{2}\sigma_k,\]
and
\[ \left[\frac{1}{2}\lambda_a, \frac{1}{2} \lambda_b\right] = i f_{ab}^{~~c}\frac{1}{2}\lambda_c.\]
These conventions lead to the wanted Standard Model gauge transformations for the component fields and the familiar relations after electroweak symmetry breaking, $e = g\sin\theta_W = g'\cos\theta_W$, where $e$ is the elementary electric charge (in natural units).

We mentioned earlier that the two Higgs superfields have opposite hypercharge. This is needed for so-called {\bf anomaly cancellation} in the MSSM. Gauge anomaly is the possibility that at loop level contributions to processes such as in Fig.~\ref{fig:anomaly} break the gauge invariance that we have established at the classical level in the Lagrangian, and ruins the predictability of the theory. This rather miraculously does not happen in the SM because it has the exactly field content it has, so that all such possible gauge anomalies exactly cancel -- we do not know of a deeper reason for why it has exactly this field content. If we have only {\it one} Higgs doublet this cancellation does not happen for the MSSM. With two Higgs doublets, and with opposite hypercharge, it does.

@@

For $SU(3)_c$ to be conserved for the Yukawa terms, we need to have colour singlets. Some of these terms are colour singlets by construction since they do not contain any coloured fields -- the $LH\bar{E}$ and $LL\bar{E}$ terms. The terms with only two quark superfields contain left-handed Weyl spinors for quarks and anti-quarks, which form $SU(3)_c$ singlets if the superfields come in colour--anti-colour pairs. In representation language the superfields (and as a consequence their component fields) are in the $\mathbf 3$ and $\mathbf{\bar 3}$ representations of $SU(3)_c$. Written with explicit  colour indices we have for example $L_i Q_j \bar{D}_k= L_i i\sigma_2 Q_j^a \bar{D}_k^a$, where $a$ is the colour index. The final term $\bar{U}_i\bar{D}_j\bar{D}_k$ is a colour singlet once we demand that it is totally anti-symmetric in the colour indices: $\bar{U}_i\bar{D}_j\bar{D}_k\equiv\epsilon_{abc}\bar{U}_i^a\bar{D}_j^b\bar{D}_k^c$. The anti-symmetry property of the Levi-Civita tensor $\epsilon_{abc}$ means that we must have $j\ne k$ in $ \lambda''_{ijk}$.

@@

The consequence of this somewhat {\it ad hoc} definition is that in all interactions the total number of incoming and outgoing supersymmetric particles must be an even number $2n$, so that the total R-number is $(-1)^{2n}=1$. This leads to the following very important phenomenological consequences:
\begin{enumerate}
\item All sparticles must be produced in pairs (ignoring the very low probability of producing four or more).
\item Sparticles must annihilate in pairs.
\item The {\bf lightest supersymmetric particle} (LSP) is absolutely stable, and every other sparticle must decay down to the LSP (possibly in multiple steps).
\end{enumerate}

>>>>

With all possible apologies, we now change notation for the component fields to what is closer to conventions in phenomenology, as opposed to pure theory. The left-handed Weyl spinors in these superfields will now be named for example $l_i$ for the spinor in the superfield $l_i$, $\nu_i$ for the one in the $\nu_i$ superfield, and $\bar e_{i}$ for the one in $\bar E_i$,  $u_{i}$ for the one in the $u_i$ superfield, $d_i$ for the one in the $d_i$ superfield, $\bar u_i$ for the one in the $\bar U_i$ superfield,  and finally $\bar d_i$ for the one in the $\bar D_i$ superfield. This means that we leave our former notation where the bar signifies right-handed Weyl spinor, but now instead signifies the left-handed Weyl spinor for the anti-particle, or, in other words, a $SU(2)_L$ singlet. If we now want to indicate a right-handed Weyl spinor we use the hermitian conjugation in (\ref{eq:Weyl_hc}) as an alternative notation, so that for example the $l_i^\dagger$ superfield contains $l_i^\dagger$ and $\bar E_i^\dagger$ contains $\bar e_i^\dagger$, and so on. The reason for this change is in part that we need a simple way to write down component field Lagrangians with fermions without greek letters and very many indices indicating particle type, as well as running out of letters since we need two superfields for one fermion.
% An alternative notation would be to write $e_{Li}$ for the left-handed Weyl-spinor in $l_i$ and $e_{Ri}$ for the left-handed Weyl-spinor in $\bar E_i$ etc.
% TODO: Work out the Weyl -> Dirac transition and re-consider the notation

@@

Looking at the construction $V\equiv qT_aV^a$ for the vector superfields in the supersymmetric Lagrangian we see that, as expected, we need one superfield $V^a$ per generator $T_a$ of the algebra, with the normal $SU(3)_c$, $SU(2)_L$ and $U(1)_Y$ vector bosons as vector component fields. We call these superfields $C^a$, $W^i$ and $B^0$, where $a=1,\ldots,8$ and $i=1,2,3$.\footnote{And there we have another W.} 
In order to be really confusing, we use the following symbols for the fermion Weyl-spinors: $\tilde g^a$, $\tilde W^+$, $\tilde W^0$, $\tilde W^-$ and $\tilde B^0$. The tilde here is supposed to tells us---just as for the scalar component fields of the scalar superfields---that they are supersymmetric partners of the known Standard Model particles. These particles are collectively known as the {\bf gauginos}.

@@

%%%%%%%%%%%%%
\section{The kinetic terms}
\label{sec:mssm_kinetic_terms}
%%%%%%%%%%%%%
It is now straight forward to write down the kinetic terms of the MSSM Lagrangian giving the matter--gauge interaction terms
\begin{eqnarray}\label{eq:kinlag}
\mathcal{L}_\text{kin} &=& L_i^\dagger e^{\frac{1}{2}g \sigma W - \frac{1}{2} g' B}L_i + Q_i^\dagger e^{\frac{1}{2}g_s\lambda C + \frac{1}{2}g\sigma W + \frac{1}{3}\cdot \frac{1}{2} g' B}Q_i \nonumber\\
&&+\bar{U}_{i}^\dagger e^{\frac{1}{2}g_s \lambda C - \frac{4}{3}\cdot \frac{1}{2} g' B}\bar{U}_i + \bar{D}_i^\dagger e^{\frac{1}{2} g_s\lambda C + \frac{2}{3}\cdot \frac{1}{2} g' B}\bar{D}_i \nonumber\\
&& + \bar{E}_i^\dagger e^{2\frac{1}{2} g' B}\bar{E}_i + H_u^\dagger e^{\frac{1}{2} g \sigma W + \frac{1}{2} g'B}H_u + H_d^\dagger e^{\frac{1}{2}g \sigma W - \frac{1}{2} g' B}H_d,
\end{eqnarray}
where $g'$, $g$ and $g_s$ are the couplings constants (strengths) of $U(1)_Y$, $SU(2)_L$, and $SU(3)_c$, respectively, and $\half\sigma_i$ and $\half\lambda_a$ are the generators of  $SU(2)_L$ and $SU(3)_c$. As a convention we assign the charges under $U(1)$, hypercharge, in units of $\frac{1}{2}g'$. For a list of the hypercharge assignments see Table~\ref{tab:hyper}.  All non-singlets of $SU(2)_L$ and $SU(3)_C$ have the same charge, the factor $\frac{1}{2}$ here is used to get by without accumulation of numerical factors since the algebras for the Pauli $\sigma_i$ ($SU(2)$) and Gell-Mann matrices $\lambda_a$ ($SU(3)$) are:
\[ \left[\frac{1}{2}\sigma_i, \frac{1}{2} \sigma_j\right] = i \epsilon_{ijk}\frac{1}{2}\sigma_k,\]
and
\[ \left[\frac{1}{2}\lambda_a, \frac{1}{2} \lambda_b\right] = i f_{ab}^{~~c}\frac{1}{2}\lambda_c.\]
These conventions lead to the wanted Standard Model gaugeinteractions for the component fields and the familiar relations after electroweak symmetry breaking, $e = g\sin\theta_W = g'\cos\theta_W$, where $e$ is the elementary electric charge (in natural units).

We mentioned earlier that the two Higgs superfields have opposite hypercharge. This is needed for so-called {\bf anomaly cancellation} in the MSSM. Gauge anomaly is the possibility that at loop level contributions to processes such as in Fig.~\ref{fig:anomaly} break the gauge invariance that we have established at the classical level in the Lagrangian, and ruins the predictability of the theory. This rather miraculously does not happen in the SM because it has the exactly field content it has, so that all such possible gauge anomalies exactly cancel---we do not know of a deeper reason for why it has exactly this field content. If we have only {\it one} Higgs doublet this cancellation does not happen for the MSSM. With two Higgs doublets, and with opposite hypercharge, it does.

@@

For $SU(3)_c$ to be conserved for the Yukawa terms, we need to have colour singlets. Some of these terms are colour singlets by construction since they do not contain any coloured fields---the $LH\bar{E}$ and $LL\bar{E}$ terms. The terms with only two quark superfields contain left-handed Weyl spinors for quarks and anti-quarks, which form $SU(3)_c$ singlets if the superfields come in colour--anti-colour pairs. In representation language the superfields (and as a consequence their component fields) are in the $\mathbf 3$ and $\mathbf{\bar 3}$ representations of $SU(3)_c$. Written with explicit  colour indices we have for example $L_i Q_j \bar{D}_k= L_i i\sigma_2 Q_j^a \bar{D}_k^a$, where $a$ is the colour index. The final term $\bar{U}_i\bar{D}_j\bar{D}_k$ is a colour singlet once we demand that it is totally anti-symmetric in the colour indices: $\bar{U}_i\bar{D}_j\bar{D}_k\equiv\epsilon_{abc}\bar{U}_i^a\bar{D}_j^b\bar{D}_k^c$. The anti-symmetry property of the Levi-Civita tensor $\epsilon_{abc}$ means that we must have $j\ne k$ in $ \lambda''_{ijk}$.

@@

The consequence of this somewhat {\it ad hoc} definition is that in all interactions the total number of incoming and outgoing supersymmetric particles must be an even number $2n$, so that the total R-number is $(-1)^{2n}=1$. This leads to the following very important phenomenological consequences:
\begin{enumerate}
\item All sparticles must be produced in pairs (ignoring the very low probability of producing four or more).
\item Sparticles must annihilate in pairs.
\item The {\bf lightest supersymmetric particle} (LSP) is absolutely stable, and every other sparticle must decay down to the LSP (possibly in multiple steps, a so-called cascade).
\end{enumerate}
